p-adic precision

### 2014-01-27 padicprec.tex
Study of SOMOS sequence (not finished)

\usepackage[plainpages=false,pdfpagelabels,colorlinks=true,citecolor=blue,hypertexnames=false]{hyperref}
\usepackage{color}
\usepackage{stmaryrd}
\usepackage{enumerate}

@@

\newcommand{\val}{\text{\rm val}}

@@

If $M$ is random square matrix over $\Zp$ of side $d$ whose every
coefficient is known up to precision $O(p^N)$, the two following
results are proved in \emph{loc. cit.}:
\begin{itemize}
\item using usual Gauss elimination and tracking $p$-adic precision 
step-by-step, the smallest precision on an entry of $L(M)$ is about
$O(p^{N - \frac{2d}{p-1}})$ on average;
\item computing $L(M)$ by evaluating Cramer-type formulas yields a
result whose every entry is known up to precision $O(p^{N - 2 \log_p 
d})$.
\end{itemize}
If $d$ is large compared to $p$, the second precision is much more
accurate than the first one. This shows once again that ``artifical''
losses of precision appears when one performs Gauss elimination by
tracking precision step-by-step.

@@

The aim of this section is to develop a new point of view on $p$-adic 
precision based on $p$-adic analysis. Actually, we shall generalize a 
bit the setting and work over a general ultrametric field (see defition
below).

@@

It may happen that natural $p$-adic ``objects'' do not lie in vector 
spaces. Actually, examples are numerous: points on the projective space 
or elliptic curves, subspaces of a fixed vector space (which lie in 
grassmannians), classes of isomorphism of certain curves (which lie in 
various moduli spaces), \emph{etc.} 
That is why we think that it is quite important to extend the formalism 
developped in Section \ref{sec:philosophy} to a more general setting. 
In this Section, we consider the quite general case of a strictly
differentiable ultrametric manifolds locally modeled on (eventually 
infinite dimensional) Banach spaces. This covers all the aforementioned 
examples.

@@

\todo{
Study SOMOS sequence
}

>>>>

\usepackage[pdfpagelabels,colorlinks=true,citecolor=blue]{hyperref}
\usepackage{color}
\usepackage{stmaryrd}
\usepackage{enumerate}
\usepackage[algoruled,vlined,french,linesnumbered]{algorithm2e}

@@

\newcommand{\val}{\text{\rm val}}
\renewcommand{\prec}{\text{\rm prec}}

@@

If $M$ is random square matrix over $\Zp$ of side $d$ whose every
coefficient is known up to precision $O(p^N)$, the two following
results are proved in \emph{loc. cit.}:
\begin{itemize}
\item using usual Gauss elimination and tracking $p$-adic precision 
step-by-step, the smallest precision on an entry of $L(M)$ is about
$O(p^{N - \frac{2d}{p-1}})$ on average;
\item computing $L(M)$ by evaluating Cramer-type formulas yields a
result whose every entry is known up to precision $O(p^{N - 2 \log_p 
d})$.
\end{itemize}
If $d$ is large compared to $p$, the second precision is much more
accurate than the first one. This shows once again that ``artifical''
losses of precision appear when one performs Gauss elimination by
tracking precision step-by-step.

@@

The aim of this section is to develop a new point of view on $p$-adic 
precision based on $p$-adic analysis. Actually, we shall generalize a 
bit the setting and work over a general ultrametric field (see definition
below).

@@

It may happen that natural $p$-adic ``objects'' do not lie in vector 
spaces. Actually, examples are numerous: points on the projective space 
or elliptic curves, subspaces of a fixed vector space (which lie in 
grassmannians), classes of isomorphism of certain curves (which lie in 
various moduli spaces), \emph{etc.} 
That is why we think that it is quite important to extend the formalism 
developped in Section \ref{sec:philosophy} to a more general setting. 
In this Section, we consider the quite general case of a strictly
differentiable manifolds locally modeled on (eventually infinite 
dimensional) ultrametric Banach spaces. This covers all the 
aforementioned examples.

@@

Recall (cf \S \ref{ssec:stepbystep}) that a SOMOS 4 sequence is a 
four-term inductive sequence defined by $u_{n+4} = \frac{u_{n+2} u_{n+4} 
+ u_{n+3}^3}{u_n}$. We recall also that SOMOS sequences exhibit the 
Laurent phenomenon: its means that the four initial terms $u_0, u_1, 
u_2, u_3$ are some indeterminates, that each $u_n$ is a Laurent 
polynomial with coefficients in $\Z$ in these indeterminates.

From now, we will always consider SOMOS sequences with values in $\Q_p$ 
(for some prime number $p$) and assume that $u_0, u_1, u_2, u_3$ are 
chosen such that all terms $u_n$'s are determined (\emph{i.e.} there is 
no division by zero). We assume also for simplicity that $u_0, u_1, u_2, 
u_3$ are all units in $\Z_p$. By Laurent phenomenon, this implies first 
that all $u_n$'s lie in $\Z_p$ and second that if $u_0, u_1, u_2, u_3$ 
are known with finite precision $O(p^N)$ then all terms $u_n$'s are 
known with the same precision.

In this last paragraph, we illustrate the theory developed in this 
article by designing an algorithm for computing the $u_n$'s at precision 
$O(p^N)$. The algorithm is presented at the top of page 
\pageref{algo:somos}. We now prove that it is correct.
%
\begin{algorithm}[t]
\SetKwInOut{Assumption}{Assumption}
\KwIn{$a,b,c,d$ --- four initial terms of a SOMOS 4 sequence $(u_n)_{n \geq 0}$}
\KwIn{$n, N$ --- two integers}
\Assumption{$a$, $b$, $c$ and $d$ lie in $\Z_p^\times$ and are known at 
precision $O(p^N)$}
\Assumption{None of $u_i$ $(0 \leq i \leq n$) is divisible by $p^N$}
\KwOut{$u_n$ at precision $O(p^N)$}
\BlankLine
$\prec$ $\leftarrow$ $N$\;
\For {$i$ from $1$ to $n-3$}{
  $\prec$ $\leftarrow$ $\prec + v_p(bd + c^2) - v_p(a)$\;
  {\bf lift} (arbitrary) 
    $b$, $c$ and $d$ to precision $O(p^{\prec})$\;
  $\prec$ $\leftarrow$ $\prec - v_p(a)$\;
  $e$ $\leftarrow$ $\frac{bd+c^2} a$; \hspace{1cm}
   \tcp{$e$ is known at precision $O(p^\prec)$}
  $a, b, c, d$ $\leftarrow$ $b + O(p^{\prec}), c + O(p^{\prec}), d + O(p^{\prec}), e$\;
}
\Return{$d + O(p^N)$}\;
\caption{\sc SOMOS$(a, b, c, d, n, N)$}\label{algo:somos}
\end{algorithm}
%
We introduce the (partial) function $f : \Q_p^4 \to \Q_p^4$ defined by 
$f(a,b,c,d) = (b,c,d,\frac{bd+c^2}a)$, so that, for all $i$, we have 
$(u_i, u_{i+1}, u_{i+2}, u_{i+3}) = f_i(u_0, u_1, u_2, u_3)$ where $f_i
= f \circ \cdots \circ f$ ($i$ times). Apparently $f$ is strictly 
differentiable on $\Q_p^\star \times \Q_p^3$ and its differential in the 
canonical basis is given by the matrix:
$$D(a,b,c,d) = \begin{pmatrix}
0 & 1 & 0 & 0 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1 \\
-\frac{bd+c^2}{a^2} & \frac d a & \frac {2c} a & \frac b a
\end{pmatrix}$$
whose determinant is $\frac{bd+c^2}{a^2}$. Thus, if the $(i+4)$-th term 
of the SOMOS sequence is defined, the mapping $f_i$ is differentiable
at $(u_0, u_1, u_2, u_3)$ and the matrix its differential at this point 
is:
$$D_i = D(u_{i-1}, u_i, u_{i+1}, u_{i+2}) \cdots
D(u_1, u_2, u_3, u_4) \cdot D(u_0, u_1, u_2, u_3).$$
%Its determinant is:
%$$\det D_i = \frac{u_{i+3}}{u_{i-1}} \cdots 
%\frac{u_5}{u_1} \cdot \frac{u_4}{u_0} =
%\frac{u_i \cdot u_{i+1} \cdot u_{i+2} \cdot u_{i+3}}
%{u_0 \cdot u_1 \cdot u_2 \cdot u_3}.$$
Moreover thanks to the Laurent phenomenon, we know that $D_i$ has
integral coefficients. Hence, letting $\varphi_i = f'_i(u_0, u_1, u_2, 
u_3)$, we get 
$$(\det D_i) \cdot p^N \Z_p^4 \subset \varphi_i(p^N \Z_p^4) 
\subset p^N \Z_p^4$$
for all integer $N$.
